Applied Measure Theory for Probabilistic Modeling

### 2022-05-12 paper/paper.tex
edits

Our work is novel in several ways. \verb|MeasureTheory.jl| has
\begin{enumerate}
    \item Explicitly represented \emph{base measures} with the same sophistication as the rest of the system, in particular more than just ``discrete or continuous'';
    \item A \emph{local} approach for determining \href{https://en.wikipedia.org/wiki/Absolute_continuity}{\emph{absolute continuity}}, which is usually a global characteristic;
    \item Multiple parameterizations for a given measure, without a proliferation of constructors; and
    \item Normalization and support constraints held separately from the data-dependent computation, allowing for greater efficiency.
\end{enumerate}

@@

The latter formulation decomposes the kernel into separate ``parameter maps'', making relationships like the one between \verb|x| and \verb|mean(κ(x))| more explicit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Densities and integrals}

\chad{I think our adoption of DensityInterface complicates this section - needs some review}

In working with measures, two computations that come up quite often are (Radon-Nikodym) \emph{differentiation} and (Lebesgue) \emph{integration}. It's often important to be able to do things like
\begin{enumerate}
    \item For measures $\mu$ and $\nu$, obtain $f=\frac{\dd\mu}{\dd\nu}$
    \item Given a measure $\nu$ and a function $f$, obtain a measure $\mu$ with $\mu(A)=\int_A f\ \dd\nu$
\end{enumerate}

Note that (1) is different from our current \verb|density| function, because no $x$ is specified. Computationally, this is called a \href{https://en.wikipedia.org/wiki/Closure_(computer_programming)}{\emph{closure}}.

In \verb|MeasureTheory|, we write these as

\begin{center}
    \verb|f = ∂(μ, ν)|\quad and\quad \verb|μ = ∫(f, ν)|\ .
\end{center}

This allows the Radon-Nikodym theorem to be expressed as
\begin{center}
    \verb|f == ∂(∫(f, ν), ν)|\quad and\quad \verb|μ == ∫(∂(μ, ν), ν)|\ .
\end{center}

Because we often we want to work instead in log-space, we also include (and tend to prefer) the functions \verb|log∂| and \verb|∫exp|, for which we would write
\begin{center}
    \verb|ℓ = log∂(μ, ν)|\quad and\quad \verb|μ = ∫exp(ℓ, ν)|\ ,
\end{center}
which follow a similar law,
\begin{center}
\verb|f == log∂(∫exp(f, ν), ν)|\quad and\quad \verb|μ == ∫exp(log∂(μ, ν), ν)|
\end{center}

Finally, in terms of \verb|density| and \verb|logdensity|, we have
\begin{verbatim}
              ∂(μ, ν)(x) == density(μ, ν, x)
           log∂(μ, ν)(x) == logdensity(μ, ν, x)
\end{verbatim}

@@

In practice, we rarely know the normalization factor $p(x)$, so we often work in terms of the unnormalized posterior density,
\[
p(\theta)\ p(x \mid \theta)\ .
\]
Because this does not include the normalization, this does not integrate to one, and so it's not a probability density function. But there are no such problems as a density for a measure; the measure defined in this way has the same base measure as that of the prior.

This \href{https://en.wikipedia.org/wiki/Pointwise_product}{\emph{pointwise product}} in probability often describes the fusion of information, in this case the fusion of information from prior and from the observations via the likelihood, but also shows up in related situation, such as \href{https://en.wikipedia.org/wiki/Sensor_fusion}{\emph{sensor fusion}} in signal processing.

More generally, this ``prior times likelihood'' can be considered as an operation that takes a measure and a function, and returns a new measure. Given a measure $\mu$ with base measure $\alpha$, and a function $f$ defined on the support of $\mu$, $\mu \odot f$ is a measure with
\[
\frac{\dd (\mu \odot f)}{\dd \alpha}(x) = \frac{\dd\mu}{\dd\alpha}(x)\ f(x)\ .
\]

In the special case where $f$ is a likelihood, it's important to keep track of whether it's represented in log space. For this, we have a \verb|Likelihood| type that ``knows'' about the measure it was derived from and adapts according to whether the user calls \verb|density| or \verb|logdensity|. The general method
\begin{verbatim}
    ⊙(::AbstractMeasure, ::Likelihood)
\end{verbatim}
defaults to constructing a \verb|PointwiseProductMeasure| to hold the prior and likelihood, but we'll be extending this with more efficient methods for cases with conjugate priors.

In the absence of conjugacy, the extra structure of this suspended form is convenient, for example to build approximations to a given measure of interest.

The above \verb|Likelihood| type is specified using a kernel and observed data. Though it is not an \verb|AbstractMeasure|, we include a \verb|logdensity| method for computational convenience. The core of the implementation is then
\begin{verbatim}
    struct Likelihood{K,X}
        k::K
        x::X
    end
\end{verbatim}

@@

For these reasons, we break the representation of the log-density into several pieces (now additive terms in log-space):
\begin{enumerate}
    \item \emph{Constant} terms,  $-\frac{1}{2}\log 2\pi$\ .
    \item \emph{Parameter-dependent} terms,  $-\log \sigma$\ .
    \item \emph{Data-dependent} terms, $-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2$\ .
\end{enumerate}

@@

\begin{itemize}
\item Borgstr\"om et al \cite{Borgstr_m_2013} describes the \verb|Fun| system in F\# in terms of \emph{measure transformer semantics}, but discusses only \emph{finite} measures.

\item Narayanan et al \cite{narayanan2016probabilistic} describe \verb|Hakaru|, a system for Bayesian modeling using measures. Here, a measure is a functional
\[
\mu[f] = \int f\ \dd\mu\ ,
\]
represented as a program. Hakaru's combinators are then expressed as compilers taking programs as the inputs.

\item Radul and Alexeev \cite{Radul2020} describe the \emph{base measure problem} of losing track of a base measure when applying a transformation, and suggest standardizing around Hausdorff measure as a solution. This problem doesn't arise for us, because the base measure is always taken into account.

>>>>

Our work is novel in several ways. \verb|MeasureTheory.jl| has
\begin{enumerate}
    \item[$\bullet\!$]Explicitly represented \emph{base measures} with the same sophistication as the rest of the system, in particular more than just ``discrete or continuous'';
    \item[$\bullet\!$]A \emph{local} approach for determining \href{https://en.wikipedia.org/wiki/Absolute_continuity}{\emph{absolute continuity}}, which is usually a global characteristic;
    \item[$\bullet\!$]Multiple parameterizations for a given measure, without a proliferation of constructors; and
    \item[$\bullet\!$]Normalization and support constraints held separately from the data-dependent computation, allowing for greater efficiency.
\end{enumerate}

@@

The latter formulation decomposes the kernel into separate ``parameter maps'', making relationships between \verb|x| and the parameters of \verb|κ(x)| more explicit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Densities and integrals}

% \chad{I think our adoption of DensityInterface complicates this section - needs some review}

% In working with measures, two computations that come up quite often are (Radon-Nikodym) \emph{differentiation} and (Lebesgue) \emph{integration}. It's often important to be able to do things like
% \begin{enumerate}
%     \item[$\bullet\!$]For measures $\mu$ and $\nu$, obtain $f=\frac{\dd\mu}{\dd\nu}$
%     \item[$\bullet\!$]Given a measure $\nu$ and a function $f$, obtain a measure $\mu$ with $\mu(A)=\int_A f\ \dd\nu$
% \end{enumerate}

% Note that (1) is different from our current \verb|density| function, because no $x$ is specified. Computationally, this is called a \href{https://en.wikipedia.org/wiki/Closure_(computer_programming)}{\emph{closure}}.

% In \verb|MeasureTheory|, we write these as

% \begin{center}
%     \verb|f = ∂(μ, ν)|\quad and\quad \verb|μ = ∫(f, ν)|\ .
% \end{center}

% This allows the Radon-Nikodym theorem to be expressed as
% \begin{center}
%     \verb|f == ∂(∫(f, ν), ν)|\quad and\quad \verb|μ == ∫(∂(μ, ν), ν)|\ .
% \end{center}

% Because we often we want to work instead in log-space, we also include (and tend to prefer) the functions \verb|log∂| and \verb|∫exp|, for which we would write
% \begin{center}
%     \verb|ℓ = log∂(μ, ν)|\quad and\quad \verb|μ = ∫exp(ℓ, ν)|\ ,
% \end{center}
% which follow a similar law,
% \begin{center}
% \verb|f == log∂(∫exp(f, ν), ν)|\quad and\quad \verb|μ == ∫exp(log∂(μ, ν), ν)|
% \end{center}

% Finally, in terms of \verb|density| and \verb|logdensity|, we have
% \begin{verbatim}
%               ∂(μ, ν)(x) == density(μ, ν, x)
%            log∂(μ, ν)(x) == logdensity(μ, ν, x)
% \end{verbatim}

@@

In practice, we rarely know the normalization factor $p(x)$, so we often work in terms of the unnormalized posterior density, and write
\[
{\underbrace{p(θ \mid x)}_\text{posterior}}
 \propto 
{\underbrace{p(\theta)}_\text{prior}}\ 
{\underbrace{p(x \mid \theta)}_\text{likelihood}} .
\]
Because the product does not include normalization, in general it does not integrate to one, and so it's not a probability density function. But there are no such problems as a density for a measure.

The \href{https://en.wikipedia.org/wiki/Pointwise_product}{\emph{pointwise product}} in probability often describes the fusion of information, in this case the fusion of information from prior and from the observations via the likelihood, but also shows up in related situation, such as \href{https://en.wikipedia.org/wiki/Sensor_fusion}{\emph{sensor fusion}} in signal processing.

Given a \emph{parameter space} $\Theta$ and \emph{observation space} $X$, a \verb|Likelihood| $\ell$ consists of a kernel $\ell_κ : \Theta \to \cM(X)$ and an observation $\ell_x \in X$. Given a measure $\rho \in \cM(X)$, $\ell$ can be treated as a function $\Theta \to \RR_{\ge 0}$, given by
\[
    \ell(\theta) \equiv \frac{\dd\ell_\kappa(\theta)}{\dd\rho}(\ell_x)
\]

In \verb|MeasureTheory|, we take $\rho$ to be the \emph{root measure} of $\ell_\kappa(\theta)$
Note that $\ell$ takes a point in $\Theta$ to an evaluated density in a different space (the observation space $X$). In particular, a likelihood is neither a measure nor a density.

Rather, a given likelihood acts on measures, taking a ``prior'' measure $\mu$ to ``unnormalized posterior'' measure $\mu \odot \ell$. Relative to another measure $\alpha$, this has density

\[
    {\underbrace{\frac{\dd(\mu\odot\ell)}{\dd\alpha}(\theta)}_\text{unnormalized posterior}}
    \equiv
    {\underbrace{\frac{\dd\mu}{\dd\alpha}(\theta)}_{\text{prior}}}\ {\underbrace{\vphantom{\frac{\dd(\mu\odot\ell)}{\dd\alpha}(\theta)}\ell(\theta)}_{\text{likelihood}}}
\]

@@

For these reasons, we break the representation of the log-density into several pieces (now additive terms in log-space):
\begin{enumerate}
    \item[$\bullet\!$]\emph{Constant} terms,  $-\frac{1}{2}\log 2\pi$\ .
    \item[$\bullet\!$]\emph{Parameter-dependent} terms,  $-\log \sigma$\ .
    \item[$\bullet\!$]\emph{Data-dependent} terms, $-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2$\ .
\end{enumerate}

@@

\begin{itemize}
\item[$\bullet\!$]Borgstr\"om et al \cite{Borgstr_m_2013} describes the \verb|Fun| system in F\# in terms of \emph{measure transformer semantics}, but discusses only \emph{finite} measures.

\item[$\bullet\!$]Narayanan et al \cite{narayanan2016probabilistic} describe \verb|Hakaru|, a system for Bayesian modeling using measures. Here, a measure is a functional
\[
\mu[f] = \int f\ \dd\mu\ ,
\]
represented as a program. Hakaru's combinators are then expressed as compilers taking programs as the inputs.

\item[$\bullet\!$]Radul and Alexeev \cite{Radul2020} describe the \emph{base measure problem} of losing track of a base measure when applying a transformation, and suggest standardizing around Hausdorff measure as a solution. This problem doesn't arise for us, because the base measure is always taken into account.
